---
title: "rENM Framework User Manual"
author: "John L. Schnase"
version: "0.1.0"
date: "2026-04-20"
number-sections: true
number-depth: 2
format:
  docx:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
  html:
    toc: true
    self-contained: true
  ipynb:
    toc: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
  results: 'hide'
  fig-show: 'hide'
---

{{< pagebreak >}}

# Introduction

The **rENM framework** is a modular suite of R packages designed to support the development and analysis of *retrospective ecological niche models* (rENMs). Retrospective ecological niche modeling integrates historical species occurrence records with time-matched environmental data to reconstruct the spatio-temporal dynamics of species’ responses to changing environmental conditions. By revealing these long-term patterns, rENMs provide a powerful observational lens for addressing biological questions and assessing both the current and future conservation status of species.

Although this analytical approach is broadly applicable across taxa, the current implementation of the rENM framework is designed to investigate climate-driven dynamics in North American bird species. The framework spans approximately 45 years (1980–2020), leveraging citizen science observations from the Cornell Lab of Ornithology’s eBird database alongside environmental data derived from NASA Earth system models.

Additional information can be found in the following publications and the project's GitHub repository at: <https://github.com/rENM-Framework>.

------------------------------------------------------------------------

**Selected Publications**

Schnase, J.L., M.L. Carroll, R.L. Gill, G.S. Tamkin, J. Li, S.L. Strong, T.P. Maxwell, M.E. Aronne, and C.S. Spradlin. Toward a Monte Carlo approach to selecting climate variables in MaxEnt. PLoS One. 2021. Mar 3;16(3):e0237208. [doi:10.1371/journal.pone.0237208](https://doi.org/10.1371/journal.pone.0237208). PMID: 33657125; PMCID: PMC7928495.

Schnase, J.L. and Carroll, M.L. 2022. Automatic variable selection inf ecological niche modeling: a case study using Cassin's Sparrow (*Peucaea cassinii*). PLoS One. 2022 Jan 21;17(1):e0257502. [doi: 10.1371/journal.pone.0257502](https://doi.org/10.1371/journal.pone.0257502). PMID: 35061658; PMCID: PMC8782318.

Schnase, J.L., and M.L. Carroll. 2023. “The MMX Toolkit: High-Performance, Reanalysis-Based Climatic Suitability Modeling to Advance Avian Conservation.” In Proceedings of the 2023 Conference on Big Data from Space (BiDS’23): 6-9 November 2023, edited by P. Soille, S. Lumnitz, and S. Albani, 299–303. Austrian Center, Vienna: Publications Office of the European. <https://doi.org/10.2760/46796.>

Schnase, J. L., M. L. Carroll, P. M. Montesano, and V. A. Seamster. 2024. Complex changes in climatic suitability for Cassin’s Sparrow (*Peucaea cassinii*) revealed by retrospective ecological niche modeling. Journal of Field Ornithology 95(1):9. <https://doi.org/10.5751/JFO-00432-950109>

Schnase, J. L., M. L. Carroll, P. M. Montesano, and V. A. Seamster. 2025. Shifts in breeding phenology for Cassin’s Sparrow (*Peucaea cassinii*) over four decades. Journal of Field Ornithology 96(3):3. <https://doi.org/10.5751/JFO-00691-960303>

Schnase, J. L., M. L. Carroll, P. M. Montesano, and V. A. Seamster. 2026. Shifts in seasonal climatic suitability for Cassin’s Sparrow (*Peucaea cassinii*) over four decades. The Southwestern Naturalist, 70(1):1-17. <https://doi.org/10.1894/0038-4909-70.1.7>

------------------------------------------------------------------------

This manual guides users through the rENM modeling and analysis workflow. Detailed documentation for individual functions is available within the respective package manuals.

The rENM framework is an active research platform under ongoing development. It is provided for educational and research purposes only and is distributed without formal technical support. Feedback and contributions are welcome.

Thank you for your interest.

**John Schnase**\
rENM.Framework\@gmail.com

{{< pagebreak >}}

# The rENM Workflow

{{< pagebreak >}}

# Configuring the Host Environment

The most straightforward way to configure your local computing environment is to complete the initial setup steps in a terminal window. The examples provided here assume a macOS environment using the Bash shell, which is the system I use.

The rENM framework relies on a simple, standardized filesystem structure. At the top level is a base project directory. Within this directory, a `data` subdirectory stores input datasets, including species occurrence records and environmental predictor variables. In this initial setup, you will install an example dataset and create a symbolic link that points to this sample collection.

Model outputs and analysis results are written to a `runs` subdirectory within the project directory, providing a clear and reproducible organization for all rENM analyses.

PROJECT_DIRECTORY/\
\|\_\_\_ data -\> (symbolic link to example_data)\
\|\_\_\_ example_data/...\
\|\_\_\_ runs/...

You only need to configure your host environment once. After completing this initial setup, you can move into the RStudio environment to finalize the rENM framework installation and begin your analyses.

In the examples that follow, I use `~/rENMtest` as the project directory for illustration. In practice, you should choose a directory name that fits your own workflow. For example, I typically use `~/rENM` in my own work

## Creating the project directory

You'll first need to create the base project directory and the `runs` subdirectory. Open a terminal window and run the following commands:

``` bash
#!/usr/bin/env bash

# set project directory name
PROJECT_DIRECTORY="$HOME/rENMtest"

# create initial directory structure
mkdir -p "$PROJECT_DIRECTORY"
mkdir -p "$PROJECT_DIRECTORY/runs"
```

## Installing the sample data set

Next, you'll want to install the example data set. Make sure your current working directory is the project base directory you just created, then run the following commands from a terminal window:

``` bash
#!/usr/bin/env bash

# set project directory name (same as above ...)
PROJECT_DIRECTORY="$HOME/rENMtest" 
cd ${PROJECT_DIRECTORY}

# download and unzip example data set
DATA_URL="https://storage.googleapis.com/renm_data/example_data.zip"
ZIP_FILE="example_data.zip"
curl -L -o "${ZIP_FILE}" "${DATA_URL}"
unzip -o "${ZIP_FILE}"

# create symbolic link
ln -sfn example_data data

# clean up the installation
rm -rf __MACOSX
rm -f "$ZIP_FILE"
```

# Configuring the RStudio Environment

From this point forward, all work is performed in the RStudio environment using R. Begin by specifying the location of the base project directory you created during the previous setup steps. This allows both RStudio and the rENM framework to locate and manage project files consistently.

## Setting the project directory location

The following command defines this location in the `.Renviron` file. This step only needs to be completed once during the initial setup. After running the command, I recommend that you leave it commented out in future sessions. Be sure that the base directory specified in this command matches the directory you created in the previous steps.

```{r}

# # set the location of the project directory
# write(
#   paste0('RENM_PROJECT_DIR="', file.path(Sys.getenv("HOME"), "rENMtest")),
#   file = "~/.Renviron",
#   append = TRUE
# )
# s
# # reload (or restart R)
# readRenviron("~/.Renviron")
# Sys.getenv("RENM_PROJECT_DIR")
```

## Installing the framework packages

Next, install the rENM framework’s collection of modular packages. These can be obtained directly from the project’s GitHub repository, as shown below. The framework currently consists of six modules. This installation step only needs to be completed once.

```{r}

# # install rENM framework packates from the GitHub repository
# install.packages("remotes")
# remotes::install_github("rENM-Framework/rENM.core")
# remotes::install_github("rENM-Framework/rENM.data")
# remotes::install_github("rENM-Framework/rENM.model")
# remotes::install_github("rENM-Framework/rENM.analysis")
# remotes::install_github("rENM-Framework/rENM.reports")
# remotes::install_github("rENM-Framework/rENM.ai")
```

## Loading the rENM framework modules

Now, load the installed rENM framework packages.

```{r}

# load the installed rENM framework packages
library(rENM.core)
library(rENM.data)
library(rENM.model)
library(rENM.analysis)
library(rENM.ai)
library(rENM.reports)
```

# Performing an rENM Analysis

Species within the rENM framework are identified using a four-character banding code (alpha code). To demonstrate the framework’s capabilities, the example dataset includes three species: Brown-capped Rosy-Finch (*Leucosticte australis*; BCRF), Cassin’s Sparrow (*Peucaea cassinii*; CASP), and Greater Roadrunner (*Geococcyx californianus*; GRRO). These species represent a range of ecological contexts, providing examples of small, medium, and large geographic distributions, as well as varying occurrence record densities.

## Viewing installed species data

Use the `show_species()` helper function to display the full list of species available in the framework’s example dataset. The four-letter alpha code (`ALPHA.CODE`) is used to identify the species selected for analysis in the next step.

```{r}

# show available species
rENM.core::show_species()
```

## Setting the species banding code

```{r}

# set four-letter banding code of a bird species
alpha_code <- "BCRF"
# alpha_code <- "CASP"
# alpha_code <- "GRRO"
```

# Data Assembly

## Occurrence data

### - Gathering eBird records

The first step in an rENM time series analysis is to prepare the input occurrence data and environmental variables. This begins with extracting species-specific occurrence records from a downloaded eBird Basic Dataset (EBD) text file located in the base data directory.

The function described below reads the raw EBD file, filters records with valid geographic coordinates, and writes a cleaned CSV file to the run directory. It then partitions the occurrence data into nine 5-year temporal bins spanning 1980–2020.

Each temporal subset is saved as a separate CSV file in the run directory under `<alpha_code>/_occs/tmp/` for use in subsequent processing steps.

```{r}

# extract ebird occurrence records from the base collection
rENM.data::get_ebird_occurrences(alpha_code)
```

### - Removing duplicate records

Next, remove duplicate records from the occurrence files in the `<alpha_code>/_occs/tmp/` run directory.

```{r}

# de-duplicate occurrence records
rENM.data::remove_duplicate_occurrences(alpha_code)
```

### - Spatial thinning

You may optionally apply spatial thinning to the occurrence records. Thinning reduces spatial clustering by enforcing a minimum distance between points, helping to mitigate sampling bias. For example, analyses of Cassin’s Sparrow typically use a thinning distance of 8 km (\~5 miles) to reduce the likelihood of counting the same individual multiple times. This step can be skipped entirely or configured by setting the `thin_distance` parameter, as shown below.

Thinning can be time-consuming. By default, temporal bins are processed sequentially. A parallel processing option is also available and can reduce runtime; however, if system resources are limited or uncertain, the default sequential approach is recommended. Comment out the option you do not use.

```{r}

# set thinning distance (radius in km)
thin_distance <- 1 

# thin records in the _occs/tmp/ run collection [sequential version]
# thin_occurrences(alpha_code, thin_distance)

# thin records in the _occs/tmp/ run collection [parallel version]
rENM.data::thin_occurrences2(alpha_code, thin_distance)
```

### - Limiting the record count

There may be times when you'd like to set an upper limit on the number of occurrence records retained within each temporal bin. This can be useful for reducing excessively high record densities, which may negatively affect model performance. This step is optional.

```{r}

# set upper bound on the number of occurrence records
record_count <- 250

# thin records to upper bound
rENM.data::limit_record_count(alpha_code, record_count)
```

### - Wrapping up occurrence data prep

We finish setting up occurrence data by moving the prepared per-bin occurrence files from the temporary *\<alpha_code\>/\_occs/tmp/* directory into the main *\<alpha_code\>/\_occs/* directory for our species run. After copying the files, the temporary directory is removed. This is the finial cleanup step before preparing environmental data.

```{r}

# move filtered records to the <alpha_code>/_occs/ run collection and tidy up
rENM.data::tidy_occurrences(alpha_code)
```

## Environmental data

### - Setting the spatial extent

Preparing environmental data for modeling involves several steps, some of which are optional. The first step is to define the spatial extent of the analysis. This step is required.

The framework provides three options for specifying model extent:\
(1) derive extent from the species’ eBird occurrence records,\
(2) derive extent from the species’ USGS GAP range, or\
(3) define the extent manually.

Select the desired approach by commenting out the options you do not intend to use. For guidance on manually specifying parameters, run `?set_extent` in the RStudio console.

```{r}

# set model extent using the usgs gap range
rENM.data::find_range_extent(alpha_code)

# set model extent using ebird occurrences records
# rENM.data::find_occurrence_extent(alpha_code)

# manually set the model extent
# rENM.data::set_extent(alpha_code)

```

### - Assembling MERRA variables

We're now ready to copy environmental variables from the base data collection into the species-specific run directory. Begin by selecting the MERRA-2 and/or MERRAclim-2 variables to include in the analysis. Use the `show_variables()` utility in the RStudio console to display the full list of available variables. Assign your selections to the `m2_vars` and `mc_vars` parameters, as shown below. Setting either parameter to `NULL` will include all available variables of that type. The example dataset includes the variables used in Schnase et al. (2026), described above. Additional details are provided in the package documentation.

```{r}

# MERRA-2 microclimatic variables
# m2_vars = c("evpintr", "evpsoil", "prevtot", "qv2m",
#             "speed", "swgnt", "tautot")

# MERRAclim-2 macroclimatic variables
# mc_vars = c("bio3", "bio5", "bio8", "bio13",
#             "bio14", "bio15", "bio18")
```

Once variables are selected, the extraction step copies them from the base collection, crops each layer to the defined spatial extent, and writes the results to the run directory under `<alpha_code>/_vars/` for subsequent processing.

```{r}

# extract merra variables and move to the run directory
rENM.data::get_merra_variables(
  alpha_code = alpha_code,
  m2_vars = NULL,
  mc_vars = NULL)
```

# Time Series Construction

## Staging occurrence records

With input data prepared, the next step is to construct the rENM time series. This begins by moving the processed occurrence records from the `<alpha_code>/_occs/` directory into a structured time series directory.

The function described below creates a `TimeSeries` directory and organizes it into nine 5-year bins labeled by year (1980–2020). Within each bin, occurrence records are stored in an `occs` subdirectory. The resulting structure is: `<alpha_code>/TimeSeries/<year>/occs/` .This organization becomes clear after running the function and inspecting the resulting `TimeSeries` directory.

```{r}

# move prepared occurrence files from the <alpha_code>/_occs/ directory 
#   to 5-year bins within the <alpha_code/TimeSeries/occs/ directory
rENM.model::stage_occurrences(alpha_code)

```

## Staging environmental variables

Environmental variables are now staged for time series analysis. Two approaches are available; select only one.

### - Option 1 - Simple staging

This function performs simple staging by transferring all cropped environmental predictors from the `<alpha_code>/_vars/` directory to the `TimeSeries` directory without modification. Files are organized into a `vars` subdirectory within each temporal bin created in the previous step. The resulting structure is: `<alpha_code>/TimeSeries/<year>/vars/` where `<year>` spans 1980–2020 in 5-year increments.

```{r}

# move all the cropped environmental variables from the <alpha_code>/_vars/
#  directory to 5-year bins within the <alpha_code/TimeSeries/vars/ directory
# rENM.model::stage_all_variables(alpha_code)
```

### - Option 2 - Staging with down selection

With this option, a Monte Carlo–based variable selection procedure (Schnase et al. 2022) is first applied to the variables in `<alpha_code>/_vars/`. This step identifies the highest-performing variables, which are then transferred to the `TimeSeries` directory for further processing.

```{r}

# down select variables (maxent version)
# rENM.model::screen_by_convergence1(alpha_code)

# down select variables (maxnet version)
rENM.model::screen_by_convergence2(alpha_code)
```

```{r}

# move the down selected, cropped environmental variables from the 
#  the <alpha_code>/_vars/ directory to 5-year bins within the 
#  <alpha_code/TimeSeries/vars/ directory
rENM.model::stage_screened_variables(alpha_code)
```

## Reducing covariance

This optional step reduces covariance among the staged environmental variables, records the results, and retains a subset of non-collinear predictors for further processing. By default, a “light-touch” approach is applied using variance inflation factors (VIF) and pairwise correlation thresholds. A relaxed VIF threshold (10) and a high correlation threshold (0.95) are used to remove only strongly redundant variables while preserving alternative representations of environmental gradients. This helps maintain the key dimensions of environmental structure identified during earlier screening steps. In practice, improved representation of environmental structure may be achieved by omitting this step.

```{r}

# reduce covariance among staged variables
# rENM.model::reduce_covariance(alpha_code)
```

## Computing the time series

With occurrence records and environmental variables staged, the next step is to generate a time series of ensemble models. The `create_timeseries()` function orchestrates this process by making parallel calls to `create_ensemble_model()` to build models for each temporal bin. By default, ensemble predictions are computed as the unweighted average of five model types: `maxnet`, random forest (`rf`), boosted regression trees (`brt`), generalized linear models (`glm`), and multivariate adaptive regression splines (`mars`). For additional details and configuration options, refer to the documentation by running `?create_ensemble_model` in the RStudio console.

```{r}

# create an rENM time series
rENM.model::create_timeseries(alpha_code)
```

# Time Series Analysis

## Suitability trend analysis

### - Finding the suitability trend

This required step computes per-cell Theil–Sen trend estimates from the time series of predicted suitability rasters and derives Mann–Kendall statistics (p-values and Z-scores) to assess temporal trends in climatic suitability. The resulting trend raster serves as the foundation for subsequent analyses.

```{r}

# compute the thiel-sen trend across the time series
rENM.analysis::find_suitability_trend(alpha_code)

```

### - Finding overall trend percentages

This is the first of several different ways of studying the suitability trends that we've derived from our rENM time series. This step provides an initial summary of the suitability trends derived from the rENM time series. The function below computes cell-wise trend sign statistics from the suitability trend raster and summarizes the proportions of positive, negative, and neutral (zero) trends *across the full modeled extent.*

```{r}

# find proportions of pos, neg, and zero values across modeled extent
rENM.analysis::find_trend_percentages(alpha_code)
```

### - Finding range trend percentages

The function below computes cell-wise trend sign statistics from the suitability trend raster and summarizes the proportions of positive, negative, and neutral (zero) trends *within the species’ USGS GAP range.*

```{r}

# find proportions of pos, neg, and zero values across USGS GAP range
rENM.analysis::find_range_change_percentages(alpha_code)
```

### - Finding state-level suitability trends

We can now perform a state-level analysis of climatic suitability trends based on the species' USGS GAP range. The function below computes summary metrics for each state intersecting the GAP range and generates a map illustrating spatial patterns alongside state-level statistics.

```{r}

# create state-level suitability trend analysis and map for a species
rENM.analysis::create_state_trend_analysis(alpha_code)
```

## Centroid trend analysis

### - Finding temporal trends

Another approach to analyzing climatic suitability trends is to examine spatial shifts in the weighted centroid of suitability over time. This function fits Bayesian linear models to annual centroid coordinates (latitude and longitude) and writes the resulting plots and summary statistics to the species’ run directory.

```{r}

# analyze temporal trends in suitability centroids
rENM.analysis::analyze_weighted_centroids(alpha_code)
```

### - Finding bioclimatic velocity

Historical centroid trajectories can also be used to quantify the direction, distance, and velocity of suitability shifts. This step computes these metrics and writes the results to the species’ run directory.

```{r}

# finds bioclimatic verlocity from weighted centroids
rENM.analysis::find_bioclimatic_velocity(alpha_code)
```

To support visualization of long-term shifts, this function generates a suitability trend map with an overlay of the centroid displacement vector. The resulting map is saved to the species’ run directory.

```{r}

# plot and save suitability trend map with centroids
rENM.analysis::save_trend_plot_with_centroids(alpha_code)
```

## Variable trend analysis

### - Gathering ranked variables

This step consolidates ranked variable importance results across the time series into a single summary dataset for the species. The output is a unified CSV file that can be used for subsequent variable analysis.

```{r}

# gather variable rankings
rENM.analysis::gather_variable_contributions(alpha_code)
```

### - Summarizing variable contributions

Using this summary, this function selects variables based on average relative contribution and generates a comprehensive set of visualizations and summary statistics.

```{r}

# summarize variable contributions
rENM.analysis::summarize_variable_contributions(alpha_code)
```

## Experimental metrics

### - Bioclimatic acceleration

The following function wraps *find_suitability_change_trend()* and *plot_suitability_change_trend()* to generate a map highlighting areas of accelerating and decelerating trends in climatic suitability. This step is experimental but provides an additional perspective on the spatial and temporal dynamics of historical climatic influences on suitability.

```{r}

# create a climatic suitability trend acceleration/deceleration map
rENM.analysis::create_suitability_change_map(alpha_code)
```

### - Hot spot analysis

This metric identifies areas where climatic suitability is declining at an increasing rate. The function generates a map highlighting these regions. This analysis is exploratory and may provide insight into areas of heightened vulnerability to climate change.

```{r}

# create a hot spot map
rENM.analysis::create_hot_spot_map(alpha_code)
```

# GenAI Analysis (In development)

## Suitability trend analysis

This experimental module uses generative AI (GenAI) to support suitability trend analysis. In the first step, the function assembles an AI-ready package containing suitability trend rasters, summary tables, species metadata, and a structured prompt. In the second step, this package is submitted to ChatGPT (GPT-5.1) for analysis. Finally, the generated results are compiled into a PDF document.

```{r}

# build a data package for submission to ChatGPT
rENM.ai::assemble_suitability_ai_package(alpha_code)

# submit the package using the OpenAI Reponses API
rENM.ai::submit_suitability_ai_package(alpha_code)

# render the returned .docx file into a .pdf file
rENM.ai::render_ai_docs(alpha_code)

```

## Environmental structure analysis

## rENM-based suitability projection

## rENM-based conservation planning

# Report Generation

The following functions compile key outputs from the preceding analysis steps to generate summary maps, tables, and reports. Because these functions depend on products created earlier in the workflow, all prior steps must be completed before proceeding to summary and report generation.

## Assemble suitability results

```{r}

# gather time series suitability maps
rENM.reports::gather_suitability_maps(alpha_code)

# gather range maps
rENM.reports::gather_range_maps(alpha_code)

# gather suitability trend and hot spot stats
rENM.reports::gather_suitability_trend_stats(alpha_code)

# create a suitability trend summary table
rENM.reports::create_suitability_trend_summary_table(alpha_code)

# create a suitability time series summary page
rENM.reports::assemble_suitability_timeseries_page(alpha_code)

# create a range time series summary page
rENM.reports::assemble_range_timeseries_page(alpha_code)

# create a suitability trends, acceleration/decleration page
rENM.reports::assemble_suitability_trends_page(alpha_code)

# create a state trends and state hot spot page
rENM.reports::assemble_state_trends_page(alpha_code)
```

## Assemble variable results

```{r}

# create variable trend summary table
rENM.reports::create_variable_trend_summary_table(alpha_code)

# create variable trends page
rENM.reports::assemble_variable_trends_page(alpha_code)

# gather trend and rate-of-change maps for the top 10 variables
rENM.reports::gather_top_variable_trend_maps(alpha_code)

# create summary page for the top 10 variables
rENM.reports::assemble_variable_trend_maps_page(alpha_code)
```

## Assemble centroid results

```{r}

# create centroid trend summary table
rENM.reports::create_centroid_trend_summary_table(alpha_code)

# create a suitability trends with centroids page
rENM.reports::assemble_centroid_trends_page(alpha_code)
```

## Assemble final report

```{r}

# create a final report comprising all summary pages
rENM.reports::assemble_final_report(alpha_code)

# these are the default pages, edit as needed ...
# rENM.reports::assemble_final_report(alpha_code, 
#                                     pages = c("Suitability-Trend-Analysis",
#                                               "Suitability-Trends",
#                                               "State-Trends",
#                                               "Centroid-Trends",
#                                               "Suitability-TimeSeries",
#                                               "Range-TimeSeries",
#                                               "Variable-Trends",
#                                               "Variable-Trend-Maps1",
#                                               "Variable-Trend-Maps2"))
```

# Automated Workflows

# APPENDICES

## Technical specifications

## Data extensions

### - Adding new species

### - Adding new variables
